Nix for Polyglot, Reproducible Data Science Workflows

### 2025-10-22 paper.qmd
Update paper.qmd

- **[rixpress]{.pkg}** orchestrates polyglot analytical pipelines using
[Nix]{.pkg} as the build engine. Each pipeline step runs in its own
hermetically sealed environment with automatic caching and dependency tracking.
This enables true polyglot workflows where [R]{.proglang}, [Python]{.proglang} and 
[Julia]{.proglang} steps coexist with different dependencies.

@@

Defining a reproducible environment with [rix]{.pkg} addresses the first major
challenge of reproducibility. The second challenge is reliably and efficiently
executing the analysis workflow within that environment, which is the role of
the [rixpress]{.pkg} package.

@@

The user defines the pipeline in an [R]{.proglang} script (`gen-pipeline.R`)
using functions inspired by [targets]{.pkg}. The underlying logic for each step
is encapsulated in separate helper scripts (`functions.jl`, `functions.py`, and
`functions.R`), the full contents of which are detailed in the Appendix. The
high-level orchestration script demonstrates how `{rixpress}` defines a
granular, multi-step machine learning workflow that crosses language boundaries:

@@

@@

While [rix]{.pkg} and [rixpress]{.pkg} provide significant steps forward, it is 
important to acknowledge current limitations. Both packages are relatively recent. 
[rixpress]{.pkg}, in particular, requires further development, especially to enhance
the ease of debugging complex pipelines. Its visualization capabilities are also still
fairly limited; for instance, they do not currently indicate outdated derivations when
a user modifies code that would necessitate a rebuild. These areas represent important 
avenues for future work to further mature the framework and improve the user experience.

@@

>>>>

- [rixpress]{.pkg} orchestrates polyglot analytical pipelines using
[Nix]{.pkg} as the build engine. Each pipeline step runs in its own
hermetically sealed environment with automatic caching and dependency tracking.
This enables true polyglot workflows where [R]{.proglang}, [Python]{.proglang} and 
[Julia]{.proglang} steps coexist with different dependencies. [Python]{.proglang} 
user can use [ryxpress]{.pkg}, a port of [rixpress]{.pkg}.

@@

Defining a reproducible environment with [rix]{.pkg} addresses the first major
challenge of reproducibility. The second challenge is reliably and efficiently
executing the analysis workflow within that environment, which is the role of
the [rixpress]{.pkg} package (or [ryxpress]{.pkg} for [Python]{.proglang} users).

@@

The user defines the pipeline in an [R]{.proglang} script (`gen-pipeline.R`)
using functions inspired by [targets]{.pkg}. The underlying logic for each step
is encapsulated in separate helper scripts (`functions.jl`, `functions.py`, and
`functions.R`). Further details are detailed step-by-step instructions are provided
in the Appendix. The high-level orchestration script demonstrates how `{rixpress}` defines a
granular, multi-step machine learning workflow that crosses language boundaries:

@@

As a sidenote: [Python]{.proglang} users that wish to use [ryxpress]{.pkg} 
would define the pipeline exactly the same way, as an [R]{.proglang} script.
[ryxpress]{.pkg} calls [R]{.proglang} and [rixpress]{.pkg} under the
hood to generate the `pipeline.nix` file and then build it.

@@

While [rix]{.pkg} and [rixpress]{.pkg} (and [ryxpress]{.pkg}) provide significant steps forward, it is 
important to acknowledge current limitations. Both packages are relatively recent. 
[rixpress]{.pkg} (and [ryxpress]{.pkg}), in particular, requires further development, especially to enhance
the ease of debugging complex pipelines. Its visualization capabilities are also still
fairly limited; for instance, they do not currently indicate outdated derivations when
a user modifies code that would necessitate a rebuild. These areas represent important 
avenues for future work to further mature the framework and improve the user experience.

@@

To build the pipeline from an interactive [Python]{.proglang} session, one would execute the following lines:

```{=tex}
\begin{CodeInput}
Python> from ryxpress import rxp_make
Python> rxp_make()
\end{CodeInput}
```
